---
title: "wwdata"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(here)
library(gtsummary)
```

## New Data

Today we will explore some epidemiologic data on surveillance of wastewater solids for infectious agents. These data come from the WastewaterScan team (see their dashboard here <https://data.wastewaterscan.org/>). This project samples wastewater solids from across most of the US (40 states, 150 treatment plants). Samples from wastewater treatment plants are sent aroun three times weekly to the Boehm lab at Stanford, where PCR for infections agents is performed, and data are displayed on a dashboard.You can learn more about the project here <https://www.wastewaterscan.org/en>.

You can even download the 2024-2025 data as a csv file from this Stanford Repository <https://purl.stanford.edu/jd535xb1898>.

## Loading this Project

Rather than doing this from scratch, you can download the whole project (and work on it) from Github.

Just

-   open R

-   Select File/New Project..

-   Select 'Version Control'

-   Select 'Git'

Then when asked for a Repository URL, paste in the following:

<https://github.com/higgi13425/wastewaterdata.git>

then Click on the the "Create Project" button

For future projects you can navigate to the project on Github, e.g. \
<https://github.com/higgi13425/wastewaterdata>

Click on the dropdown arrow on the green Code button, and clone the HTTPS url, by clicking on the copy button to the right.

Now you have your own local copy of the project.

This opens on the main branch. In order to avoid stomping on other people's code, create a new working branch of your own, by clicking on the branch icon (two purple branches) within the Git pane at top right. Name your working branch something like "working_myname". You can also name it after what you are working on in that particular branch, like "table_one", or "figure_3". When your branch is finished, you can create a pull request (aka PR) to ask the owner of the repository (repo) to merge your changes into the main (aka production) branch.

A quick video on pull requests (using the git command line interface or CLI) for beginners can be found in this [video](https://www.youtube.com/watch?v=nCKdihvneS0).

## Exploring the repository

You should now have a branch of the repo named something like 'working_bob'. Therre should be some infrastructure files that start with a period (.git, .gitignore, .Rproj.user) and a wastewater_files folder. There is also a wastewater.Rproj file that identifies the root level of the repo (where the here function path goes).

You don't need to worry about these unless you want to add something to the .gitignore file. You might want to do this if you have files that you don't want backed up or shared publicly, like files with protected health information (PHI). You want to make sure that any PHI files in the repo are listed in the .gitignore so that they are never backed up to git or to GitHub.

The three useful files are wastewater.qmd, a quarto file we will use to write up results; wastewater.html - the rendered version of wastewater.qmd - so that you can publish pretty results to the web; and the data file 'WWSCAN_AllPlants_SDR_2025.csv' - this is a csv file with the data that are presented in the fancy dashboard here <https://data.wastewaterscan.org/>.

## Digging into the Quarto file

Open the wastewater.qmd file. This will look familiar, as it is the origin for this HTML page. You can even scroll down to this header (Digging into the Quarto file).

We will start by

-   Reading in the csv file

-   Exploring the dataset

-   Making Table1 like tables using gtsummary

-   Looking at trends over time with ggplot2

-   Diving into Ann Arbor data to look at seasonality of infections

-   Testing a few hypotheses of your own

## Reading in the csv file

Start by running this code chunk to assign the data to the object `ww`.

```{r}
ww<- readr::read_csv('WWSCAN_AllPlants_SDR_2025.csv')
```

This `ww` object should now show up in your Environment with 28217 observations of 69 variables. Use the View() function to take a look at it. Each observation starts with a site and a date of collection, then gives a variety of results for different (abbreviated) infectious agents. There is a point estimate and a lower and upper confidence bound for each infectious agent (3 columns each). Compare the infectious agents shown on the dashboard to the column names in the ww object. Can you pick out Norovirus G2, Rotavirus, RSV, Enterovirus D68, Influenza A and B, SARS-CoV-2, Hepatitis A, Candida auris, etc. Remember that there will be 3 columns for each infectious agent, the estimate, the LC bound (LCI) and the UC bound (UCI).

## Filtering down to a Single State

Run the code below to filter the data to a single state using the State_Abbr variable.

Be sure to have already loaded the tidyverse and here packages in the initial code chunk in this qmd document.

```{r}
ww |> 
  filter(State_Abbr == 'FL') |> 
  distinct(City)
```

By adding the distinct function we can see that there are 10 different sites in Florida that participate in the Wastewater scan project. Try using distinct(Plant) or distinct(City) to get a sense of the locations of these sites.

Try this with a few other states (note that 10 states do not participate and will have no sites).

## Challenges

Try these out with your R skills

Add an R chunk to this qmd in your branch and try to answer each question.

-   How many sites are participating overall?

-   How many sites are participating in your home state?

-   Where are these sites located?

-   How many samples have been collected in your state?

-   Which site has collected the most samples? (overall, or in your state)

-   What is the largest number of samples collected by one site?

```{r}
ww |> 
  #vfilter(State_Abbr == 'CA') |> 
  group_by(Plant) |> 
  count() |> 
  arrange(desc(n))
```

## Making a Table of Sites

Use gtsummary to make a pretty HTML table of sites in one state, arranged by number of samples collected. Try to pick different states.

## Making a Table of Infectious Agents

Use gtsummary to make a pretty HTML table of the infectious agents being tracked, arranged by number of non-zero / non-missing samples collected.

## Tracking Trends

Use the select function and ggplot 2 line graphs to track time trends in one particular infection. Pick a seasonally trendy one like

-   RSV

-   Influenza A

-   COVID

-   Influenza B

-   Norovirus

-   Enterovirus D68

When are the seasonal peaks nationally?

## Go local

Challenges

-   Look at your local state or local Plant. Filter the results down to your local wastewater. Look at all of the infections over the past year. Plot incidence (Y) vs time (X axis). What seasonal trends do you see locally?

-   If you see seasonal trends in one infectious agent, are these pretty much the same across the state, or do they vary regionally? Does RSV (or Noro, or other bugs) arrive later in some regions?

## Share what You Have Learned

Contribute to the repository

-   first, save your qmd once it is working

-   render your QMD to HTML (the rightward blue arrow) to see a pretty version

-   Commit your changes (in the Git tab) - add a commit message to explain what you did.

-   Push (the upward green arrow) your branch changes to GitHub

## Is it Ready for Production?

Then make a pull request. Go to the Github site at <https://github.com/higgi13425/wastewaterdata>

If you have successfully pushed your branch changes, you will see a big green button - Compare & Pull Request. Click on this button.

-   add an informative title

-   add a detailed description of what you created in this branch

-   then scroll down to the green button - and click on Create pull request

Then I (the owner of the repo) will get this pull request, and I can compare the differences between the main branch and your branch. I can merge your changes (pull them in) to the main branch. We can collaborate by working on different tables, figures, models, etc. in distinct branches. We can also review each other's code by reviewing different branches or merges, all via GitHub.
